---
title: "Data Transformation"
author: "Pongsun B."
date: "2024-09-05"
output:
  html_document: default
  pdf_document: default
header-includes:
  - \pagenumbering{gobble}
---

## ==== First things first ====

Set working directory and call packages.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Call required packages, ex. "readxl", "tidyverse"
if(!require(readxl)){install.packages("readxl")}
if(!require(tidyverse)){install.packages("tidyverse")}
if(!require(RCurl)){install.packages("RCurl")}

```

The concepts and examples in this chapter are drawn from Mangiafico (2016), Parker (2020), Cornelissen (2022), Johnson, D. (2022), and Ryu (2022). For further details please refer to the mentioned literature.

## Defining a Function

Create the function by using the `function` keyword followed by a list of parameter names and then the function body.

```{r function, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# First function: square 
square <- function(x) { 
  return(x*x); 
  } 

cat("The square of 3 is", square(3), "\n")

# Second function: cube with a default value
cube <- function(x=5) { 
  return(x*x*x); 
  } 

cat("Calling cube with 2 :", cube(2), "\n") # will give 2^3 
cat("Calling cube :", cube(), "\n") # will default to 5^3.

# Functions that return multiple objects
powerful <- function(x) { 
  return(list(x2 = x*x, x3 = x*x*x, x4 = x*x*x*x));
  }
cat("Showing powers of 3 --", "\n")
powerful(3)

```

## Applying a Function

While traditional programming languages use loops, R has traditionally encouraged using vectorized operations and the `apply` family of functions to crunch data in batches, greatly streamlining the calculations. There is nothing to prevent you from writing loops in R that break your data into whatever chunks you want and then do an operation on each chunk. However, using vectorized functions can, in many cases, increase the speed, readability, and maintainability of your code.

```{r Loop and vectorization, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# EXAMPLE 1: sin(x) for a vector 
# Suppose you have a vector x 
x = c(0.1,0.6,1.0,1.5) 

# The bad way 
n = length(x) 
r = numeric(n) 
for (i in 1:n) { 
  r[i] = sin(x[i]) 
  } 
r

# The good way -- don't use loops 
sin(x)

# EXAMPLE 2: Compute the mean of every row of a matrix
# Let setup a matrix
N = 4
M = 10
r = matrix(runif(N*M), N, M)

# Method 1 uses loops: 
mean1 = numeric(M) 
for (i in 1:M) { 
  mean1[i] = mean(r[,i]) 
  }
mean1

# Alternatively, just say: 
mean2 = rep(1/N, N) %*% r 
mean2

# Check whether the two answers are the same
all.equal(mean1,mean2[,])

```

In recent history, the *tidyverse*, specifically the `purrr` and `dplyr` packages, has introduced new idioms into R that make these concepts easier to learn and slightly more consistent. The name purrr comes from a play on the phrase “Pure R.” A “pure function” is a function where the result of the function is determined only by its inputs, and which does not produce any side effects. This is not a functional programming concept you need to understand in order to get great value from `purrr`, however. All most users need to know is that `purrr` contains functions to help us operate “chunk by chunk” on our data in a way that meshes well with other *tidyverse* packages such as `dplyr`.

Base R has many `apply` functions—`apply, lapply, sapply, tapply, and mapply`—and their cousins, `by` and `split`. These are solid functions that have been workhorses in Base R for years. However, many users have struggled a bit with how much to focus on the Base R `apply` functions and how much to focus on the newer “tidy” approach. Therefore, we’ve chosen to try to illustrate the `purrr` approach and to acknowledge Base R approaches and, in a few places, to illustrate both. The interface to `purrr` and `dplyr` is very clean and, we believe, in most cases, more intuitive.

## Applying a Function to Each List Element

#### Problem

You have a list, and you want to apply a function to [each element of the list]{.underline}.

#### Solution 1

Use either the `lapply` function or the `sapply` function, depending upon the desired form of the result. `lapply` always returns the results in list, where as `sapply` returns the results in a vector if that is possible.

```         
lst <- lapply(lst, fun)
vec <- sapply(lst, fun)
```

These functions will call your function (`fun`, in the solution example) once for every element on your list. Your function should expect one argument, an element from the list. The `lapply` and `sapply` functions will collect the returned values. `lapply` collects them into a list and returns the list.

The “`s`” in “`sapply`” stands for “simplify.” The function tries to simplify the results into a vector or matrix. For that to happen, all the returned values must have the same length. If that length is 1 then you get a vector; otherwise, you get a matrix. If the lengths vary, simplification is impossible and you get a list.

#### `lapply` with multiple arguments

It should be noted that if the function you are passing to the `fun` argument has addition arguments [you can pass them after the function, using a comma]{.underline}.

#### `lapply` with a custom function

You can also apply a custom function with `lapply`. For that purpose you can create a function and pass its name to the `fun` argument of just write it inside the `lapply` function.

#### Solution 2

We can use `map` to apply the function to every element of a list:

```         
library(tidyverse)

lst %>%
  map(fun)
```

The `map` function will call your function once for every element on your list. Your function should expect one argument, an element from the list. The `map` functions will collect the returned values and return them in a list.

The `purrr` package contains a whole family of `map` functions that take a list or a vector and then return an object with the same number of elements as the input. The type of object they return varies based on which `map` function is used. See the help file for `map` for a complete list, but a few of the most common are as follows:

`map()` always returns a list, and the elements of the list may be of different types. This is quite similar to the Base R function `lapply`.

`map_chr()` returns a character vector.

`map_int()` returns an integer vector.

`map_dbl()` returns a floating-point numeric vector.

The Base R `lapply` function acts very much like `map`. The Base R `sapply` function is more like the other `map` functions we discussed previously in that the function tries to simplify the results into a vector or matrix.

#### Practice 1: Use `sapply` and `lapply`

Let’s say I teach an introductory statistics class four times and administer comparable final exams each time.

Each semester starts with 40 students but, not everyone makes it to the finish line; hence each semester has a different number of scores.

If the called function returns a vector, `sapply` will form the results into a matrix.

```{r lapply_sapply, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
# Create the exam scores from the four semesters
stat_score <- list(S1 = c(89, 85, 85, 86, 88, 89, 86, 82, 96, 85, 
                          93, 91, 98, 87, 94, 77, 87, 98, 85, 89, 
                          95, 85, 93, 93, 97, 71, 97, 93, 75, 68, 
                          98, 95, 79, 94, 98, 95), 
                   S2 = c(60, 98, 94, 95, 99, 97, 100, 73, 93, 91, 
                          98, 86, 66, 83, 77, 97, 91, 93, 71, 91,
                          95, 100, 72, 96, 91, 76, 100, 97, 99, 95,
                          97, 77, 94, 99, 88, 100, 94, 93, 86),
                   S3 = c(95, 86, 90, 90, 75, 83, 96, 85, 83, 84, 
                          81, 98, 77, 94, 84, 89, 93, 99, 91, 77,
                          95, 90, 91, 87, 85, 76, 99, 99, 97, 97, 
                          97, 77, 93, 96, 90, 87, 97, 88),
                   S4 = c(67, 93, 63, 83, 87, 97, 96, 92, 93, 96, 
                          87, 90, 94, 90, 82, 91, 85, 93, 83, 90,
                          87, 99, 94, 88, 90, 72, 81, 93, 93, 94,
                          97, 89, 96, 95, 82, 97))

# Count the number of students in each semester with the length function
# lapply will return a list of lengths

# sapply will return a vector of lengths

# Determine the mean and standard deviation of the scores

# Use the range function, which returns a two-element vector

# lapply with multiple arguments

# lapply with a custom function


```

#### Practice 2: Use `map` 

```{r map, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

lst <- list(
  a = c(1,2,3),
  b = c(4,5,6)
)

# Determine the average of all the numbers in each element of a list


# Create a function that could result in a character or an integer result
myfun1 <- function(x) {
  if (x > 1) {
    1
  } else {
    "Less Than 1"
  }
}

myfun1(5)
myfun1(0.5)

# Create a list of elements that we can map "myfun" to
mylst <- list(.5, 1.5, .9, 2)

# Look at how some of the "map" variants behave

# map_chr will produce a character vector and coerce the numbers into characters

# map_dbl will try to coerce a character string into a double and die trying.


```

## Applying a Function to Every Row of a Matrix

#### Problem

You have a matrix. You want to apply a function to [every row]{.underline}, calculating the function result for each row.

#### Solution

Use the `apply` function. Set the second argument to 1 to indicate row-by-row application of a function.

```         
results <- apply(mat, 1, fun) # mat is a matrix, fun is a function
```

The `apply` function will call `fun` once for each row, assemble the returned values into a vector, and then return that vector.

The function being called (`fun`, described previously) should expect one argument, a vector, which will be one row from the matrix. The function can return a scalar or a vector.

You can employ this recipe on data frames as well. It works if the data frame is homogeneous—either all numbers or all character strings. When the data frame has columns of different types, extracting vectors from the rows isn’t sensible because vectors must be homogeneous.

#### Practice

Suppose your matrix `long` is longitudinal data. Each row contains data for one subject, and the columns contain the repeated observations over time. You could calculate the average observation for each subject by applying the mean function to the rows.

```{r apply, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
# Create a longitudinal data

long_vec <- c(-1.8501520, -1.406571, -1.0104817, -3.7170704, -0.2804896,
               0.9496313,  1.346517, -0.1580926,  1.6272786, 2.4483321,
              -0.5407272, -1.708678, -0.3480616, -0.2757667, -1.2177024)

long_mat <- matrix(long_vec, byrow = T, nrow = 3)

colnames(long_mat) <- c("trial1", "trial2", "trial3", "trial4", "trial5")
rownames(long_mat) <- c("Moe", "Larry", "Curly")

# Calculate the average observation for each subject by applying the mean function to the rows


# Note that apply uses the rownames from your matrix to identify the elements of the resulting vector, which is handy

# Determine the minimum and the maximum with the range function


```

## Applying a Function to Every Row of a Data Frame

#### Problem

You have a function and you want to apply it to [every row]{.underline} in a data frame.

#### Solution

The `mutate` function will create a new variable based on a vector of values. But if we are using a function that can’t take in a vector and output a vector, then we have to do a row by row operation using `rowwise`.

We can use `rowwise` in a pipe chain to tell `dplyr` to do all following commands row-by-row:

```         
df %>%
  rowwise() %>%
  row_by_row_function()
```

```{r rowwise, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
# Create a function and apply it row by row to a data frame:
# Our function will simply calculate the sum of a sequence from a to b by c:

myfun2 <- function(a, b, c) {
  sum(seq(a, b, c))
}

# Create some data to apply this function
df <- data.frame(mn = c(1, 2, 3),
                 mx = c(8, 13, 18),
                 rng = c(1, 2, 3))

# Then, use "rowwise" to apply myfun


```

## Applying a Function to Every Column

#### Problem

You have a matrix or data frame, and you want to apply a function to [every column]{.underline}.

#### Solution

For a matrix, use the `apply` function. Set the second argument to 2, which indicates column-by-column application of the function.

```         
results <- apply(mat, 2, fun) # mat is a matrix, fun is a function
```

The `apply` function is intended for processing a matrix. Previously, we used `apply` to process the rows of a matrix. This is the same situation, but now we are processing the columns. The second argument of `apply` determines the direction:

-   1 means process row [by row]{.underline}.

-   2 means process column [by column]{.underline}.

This is more mnemonic than it looks. We speak of matrices in “rows and columns”,so rows are first and columns second; 1 and 2, respectively.

For a data frame, use the `map_df` function from `purrr`:

```         
df2 <- map_df(df, fun) # Returns a data.frame
```

You can also use `apply` on data frames, but only if the data frame is homogeneous (i.e., either all numeric values or all character strings).

The function `fun` should expect one argument: a column from the data frame.

A data frame is a more complicated data structure than a matrix, so there are more options. You can simply use `apply`, in which case R will convert your data frame to a matrix and then apply your function. That will work if your data frame contains only one type of data but will probably not do what you want if some columns are numeric and some are character. In that case, R will force all columns to have identical types, likely performing an unwanted conversion as a result.

Fortunately, there are multiple alternatives. Recall that a data frame is a kind of list: it is a list of the columns of the data frame. `purrr` has a whole family of `map` functions that return different types of objects. Of particular interest here is `map_df`, which returns a `data.frame`, thus the `df` in the name.

```{r apply and map_df, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
# Create a matrix 
mymat <- matrix(c(1, 3, 2, 5, 4, 6), nrow = 2)
colnames(mymat) <- c("t1", "t2", "t3")
mymat

# Apply the mean function to compute the mean of every column


# Load "Batch" data
load("./Data/batches.rdata")
head(batches) # The column of this data frame, at quick glance, seems to contain numbers.

# Use "map_df" to check the class of each column 


```

## Applying a Function to Parallel Vectors or Lists

#### Problem

You have a function that takes multiple arguments. You want to apply the function element-wise to vectors and obtain a vector result. Unfortunately, the function is not vectorized; that is, it works on scalars but not on vectors.

#### Solution

Use one of the `map` or `pmap` functions from the *tidyverse* core package `purrr`. The most general solution is to put your vectors in a list, then use `pmap`.

```         
lst <- list(v1, v2, v3)
pmap(lst, fun)
```

`pmap` will take the elements of `lst` and pass them as the inputs to `fun`.

If you only have two vectors you are passing as inputs to your function, the `map2_*` family of functions is convenient and saves you the step of putting your vectors in a list first. `map2` will return a list, while the typed variants (`map2_chr`, `map2_dbl`, etc.) return vectors of the type their name implies:

```         
map2(v1, v2, fun)
```

or if `fun` returns only a double, then use the typed variant of `map2`:

```         
map2_dbl(v1, v2, fun)
```

The typed variants in `purrr` functions refer to the *output* type expected from the function. All the typed variants return vectors of their respective type, while the untyped variants return lists, which allow mixing of types.

The basic operators of R, such as *x* + *y*, are vectorized; this means that they compute their result element by element and return a vector of results. Also, many R functions are vectorized.

Not all functions are vectorized, however, and those that are not typed work only on scalars. Using vector arguments produces errors at best and meaningless results at worst. In such cases, the `map` functions from `purrr` can effectively vectorize the function for you.

`map2` returns a list of lists. If we wanted the output in a vector, we could use `unlist` on the result, or use one of the typed variants, such as `map2_dbl`.

```{r map2 and pmap, warning=FALSE, list(width.cutoff=80), tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# Consider the gcd function, which takes two arguments:
gcd <- function(a, b) {
  if (b == 0) {
    return(a)
  } else {
    return(gcd(b, a %% b))
  }
}

# If we apply "gcd" to two vectors, the result is wrong answers and a pile of error messages:
# gcd(c(1, 2, 3), c(9, 6, 3))

a <- c(1, 2, 3)
b <- c(9, 6, 3)

# Use "map2" to vectorize it. This gives the element-wise GCDs between two vectors.


# Use unlist on the result to have the output in a vector 


```

The `map` family of `purrr` functions give you a series of variations that return specific types of output. The suffixes on the function names communicate the type of *vector* they will return. While `map` and `map2` return lists, since the type-specific variants are returning objects guaranteed to be the same type, they can be put in atomic vectors. For example, we could use the `map_chr` function to ask R to coerce the results into character output or `map2_dbl` to ensure the results are doubles.

```{r map2_dbl, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
# Use map2_chr and map2_dbl


# If data has more than two vectors, or the data is already in a list, we can use the "pmap" family of functions, which take a list as an input.



```

With the `purrr` functions, remember that the `pmap` family are parallel mappers that take in a *list* as inputs, while `map2` functions take two, and only two, *vectors* as inputs.

## Applying a Function to Groups of Data

#### Problem

Your data elements occur in groups. You want to process the data by groups—for example, summing by group or averaging by group.

#### Solution

The easiest way to do grouping is with the `dplyr` function `group_by` in conjunction with `summarise`. If our data frame is `df` and has a variable we want to group by named `grouping_var`, and we want to apply the function `fun` to all the combinations of `v1` and `v2`, we can do that with `group_by`.

```         
df %>%
  group_by(v1, v2) %>%
  summarize(
    result_var = fun(value_var)
  )
```

#### Practice

Let’s look at a specific example where our input data frame, `df`, contains a variable, `my_group`, which we want to group by, and a field named `values`, which we would like to calculate some statistics on:

```{r group_by, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
df2 <- data.frame(
  my_group = c("A","B","A","B","A","B"),
  values = 1:6
)



```

If you are grouping by several variables, please be aware that `summarize` will change your grouping. Each grouping becomes a single row; at the same time, it also removes the *last* grouping variable. In other words, if you group your data by `A`, `B`, and `C` and then `summarize` it, the resulting data frame is grouped only by `A` and `B`. This is surprising but necessary. If `summarize` kept the `C` grouping, each “group” would contain exactly one row, which would be pointless.

## Creating a New Column Based on Some Condition

#### Problem

You want to create a new column in a data frame based on some condition.

#### Solution

Using the `dplyr` in *tidyverse* package, we can create new data frame columns using `mutate` and then use `case_when` to implement conditional logic.

```         
df %>%
  mutate(
    new_field = case_when(my_field == "something" ~ "result", 
                          my_field != "something else" ~ "other result", 
                          TRUE ~ "all other results")
  )
```

#### Practice

The `case_when` function from `dplyr` is analogous to `CASE WHEN` in SQL or nested `IF` statements in Excel. The function tests every element and when it finds a condition that is true, it returns the value on the right side of the `~` (tilde).

The condition goes on the left of the `~`, while the resulting return value goes on the right. Each condition is separated by commas. `case_when` will evaluate each condition sequentially and stop evaluating as soon as one of the criteria returns `TRUE`. Our last line is our “or else” statement. By setting the criteria to `TRUE`, no matter what, this condition will be met if no condition above it has returned `TRUE`.

Let’s look at an example where we want to add a text field that describes a value. First let’s set up some simple example data in a data frame with one column named `vals`. Then, let’s implement logic that creates a field called `new_vals`. If `vals` is less than or equal to 2, we’ll return `2 or less`; if the value is between greater than 2 and less than or equal to 4, we’ll return `2 to 4`; and otherwise we’ll return `over 4`:

```{r case_when, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
df3 <- data.frame(vals = 1:5)

# Create a new column with mutate and case_when

```

End-of-File\
Pongsun B.\
2024-09-05
